Semi-Supervised Factored Logistic Regression for High-Dimensional Neuroimaging Data

### 2015-06-04 paper_nips/paper.tex
interm

\paragraph{Optimization.}
The common objective was optimized
by gradient descent in the SS-F LogReg parameters.
%
The required gradients were obtained by using the chain rule to
backpropagate error derivatives.
We chose the \textit{rmsprop} solver \cite{rmsprop},
a refinement of stochastic gradient descent.
\textit{Rmsprop} dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
The batch size was set to $100$
(much expected redundancy in $X_{rest}$ and $X_{task}$),
matrix parameters were initalized by Gaussian random values multiplied
by $0.004$ (i.e., gain), and
bias parameters were initalized to zero.
%
The normalization factor and the update rule for $\theta$
are given by
% XXX Michael: It is crucial, but nowhere mentioned, that the optimization is
% XXX done in an alternating manner. The whole structure relies on this, by
% XXX construction it is vulnerable to optimizing too much towards labels or
% XXX rest at a time. I also identify this as one of the major weak points
% XXX of the method, so it needs to be stated in a way that is not easily
% XXX attackable.
% XXX Maybe we could also talk less about RMSprop.
%
% XXX Danilo: I think we are not alternating, Xrest and Xtask go in
% simultaneously at every bach in every epoch
\begin{eqnarray}
  \begin{split}
    \mathbf{v^{(t+1)}} &= \rho \mathbf{v^{(t+1)}} + (1 - \rho)\left(\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})\right)^2
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})}{\sqrt{\mathbf{v^{(t+1)}} + \epsilon}},
  \end{split}
\end{eqnarray}
% XXX Michael: Why, in the formula, do we have two different notations for
% XXX partial derivatives? \partial and \nabla
% XXX Michael: Probably remove the explanations following and just state values
% XXX: Bertrand: what is f ?? Danilo: replaced f by more detailed notation
where $f$ is the loss function computed on a minibatch sample at timestep $t$,
$\alpha$ is the learning rate ($0.00001$),
$\epsilon$ a global damping factor ($10^{-6}$),
and
$\rho$ the decay rate ($0.9$ to deemphasize the magnitude of the gradient).
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
\textit{rmsprop} converged faster and with
similar or higher generalization performance.

@@

SS-F LogReg was subsequently trained ($500$ epochs) across parameter choices

@@

higher bias and smaller variance then SS-F LogReg with 100 latent directions.

@@

\begin{table}[h]
  \centering
  \resizebox{1.00\textwidth}{!}{%
\tabcolsep=0.05cm
\begin{tabular}{l|ccccc|ccccc|ccccc}
  \hline
                            & \multicolumn{5}{c|}{\textit{n = 5}}                                                                                                                                                 & \multicolumn{5}{c|}{\textit{n = 20}}                                                                                                                                                      & \multicolumn{5}{c}{\textit{n = 100}}                                                                           \\ \hline
  \textit{}                 & \multicolumn{1}{l}{$\lambda=0$}    & \multicolumn{1}{l}{$\lambda=0.25$} & \multicolumn{1}{l}{$\lambda=0.5$} & \multicolumn{1}{l}{$\lambda=0.75$} & \multicolumn{1}{l|}{$\lambda=1$} & \multicolumn{1}{l}{$\lambda=0$}  & \multicolumn{1}{l}{$\lambda=0.25$}   & \multicolumn{1}{l}{$\lambda=0.5$}   & \multicolumn{1}{l}{$\lambda=0.75$}   & \multicolumn{1}{l|}{$\lambda=1$}   & $\lambda=0$      & $\lambda=0.25$   & $\lambda=0.5$    & $\lambda=0.75$   & \multicolumn{1}{l}{$\lambda=1$}    \\ \hline
  Out-of-sample\\accuracy   & \textit{6,0\%}                     & 88,9\%                             & 95,1\%                            & \textbf{96,5\%}                    & 95,7\%                           & \textit{5,5\%}                   & 97,4\%                               & \textbf{97,8\%}                     & 97,3\%                               & 97,3\%                             & \textit{6,1\%}   & 97,2\%           & 97,0\%           & \textbf{97,8\%}  & 97,4\%                             \\
  Precision (mean)          & \textit{5,9\%}                     & 87,0\%                             & 94,9\%                            & \textbf{96,3\%}                    & 95,4\%                           & \textit{5,1\%}                   & \textbf{97,4\%}                      & 97,1\%                              & 97,0\%                               & 97,0\%                             & \textit{5,9\%}   & 96,9\%           & 96,5\%           & \textbf{97,5\%}  & 96,9\%                             \\
  Recall (mean)             & \textit{5,6\%}                     & 88,3\%                             & 95,2\%                            & \textbf{96,6\%}                    & 95,7\%                           & \textit{4,6\%}                   & \textbf{97,5\%}                      & 97,5\%                              & 97,4\%                               & 97,4\%                             & \textit{7,2\%}   & 97,2\%           & 97,2\%           & \textbf{97,9\%}  & 97,4\%                             \\
  F1 score (mean)           & \textit{4,1\%}                     & 86,6\%                             & 94,9\%                            & \textbf{96,4\%}                    & 95,4\%                           & \textit{3,8\%}                   & \textbf{97,4\%}                      & 97,2\%                              & 97,1\%                               & 97,1\%                             & \textit{5,3\%}   & 97,0\%           & 96,7\%           & \textbf{97,7\%}  & 97,2\%                             \\ \hline
  Out-of-study\\accuracy    & \textit{39,4\%}                    & 60,8\%                             & 54,3\%                            & 60,7\%                             & \textbf{62,9\%}                  & \textit{77,0\%}                  & 79,7\%                               & \textbf{81,9\%}                     & 79,7\%                               & 79,4\%                             & 79,2\%           & \textbf{82,2\%}  & 81,7\%           & 81,3\%           & \textit{75,8\%}                    \\ \hline
  \end{tabular}
}
\vspace{-0.2cm}
 \caption{\textbf{Performance of SS-F LogReg across model parameter choices.}
 Chance is at 5,6\%.
 }
  \label{table_two}
\end{table}

@@

\paragraph{Feature identification.}
We finally examined whether the models
were fit for purpose
(Figure \ref{fig_weights}).
%
To this end, we computed Pearson's correlation between the classifier weights
and the averaged neural activity map for each of the 18 tasks.
%
Ordinary logistic regression thus yielded a mean correlation
of $\rho=0.28$.
%
For SS-F LogReg ($\lambda=0.25, 0.5, 0.75, 1$),
a per-class-weight map was computed by matrix
multiplication of the two inner layers.
Feature identification thus ranged
between $\rho=0.35$ and $\rho=0.55$ for $n=5$,
between $\rho=0.59$ and $\rho=0.69$ for $n=20$, and
between $\rho=0.58$ and $\rho=0.69$ for $n=100$.
%
Consequently,
SS-F LogReg puts higher absolute weights on relevant structure.
This reflects a higher signal-to-noise ratio, in part explained
by the more BOLD-typical local contiguity.
%
Conversely, SS-F LogReg puts lower probability mass on irrelevant structure.
Despite lower interpretability, the
salt-and-pepper-like weight maps obtained from
ordinary logistic regression are sufficient for
good classification performance.
%
\begin{figure}
\begin{centering}
\includegraphics[width=0.98\textwidth]{figures/figure_weights_perc75.png}
\end{centering}
\vspace{-0.1cm}
\caption{\textbf{Classification weight maps}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} multinomial logistic regression (same
implementation but without bottleneck or autoencoder),
\textit{middle column:} semi-supervised factored logistic regression
($n=20$ latent components, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task.
Semi-supervised factored logistic regression
a) puts higher absolute weights on relevant structure,
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity (without enforcing a spatial prior).
All values are z-scored and thresholded at the $75^{th}$ percentile.
}
\label{fig_weights}
\end{figure}
%
Hence, SS-F LogReg yielded class weights 
that were much more similar to features of the respective training samples
for any choice of $\lambda$.
SS-F LogReg therefore
captures genuine properties of neural activity patterns,
rather than participant- or study-specific artefacts.

>>>>

\paragraph{Optimization.}
The common objective was optimized
by gradient descent in the SS-FLogReg parameters.
%
The required gradients were obtained by using the chain rule to
backpropagate error derivatives.
We chose the \textit{rmsprop} solver \cite{rmsprop},
a refinement of stochastic gradient descent.
\textit{Rmsprop} dictates an adaptive learning rate
for each model parameter by
scaled gradients from a running average.
The batch size was set to $100$
(much expected redundancy in $X_{rest}$ and $X_{task}$),
matrix parameters were initalized by Gaussian random values multiplied
by $0.004$ (i.e., gain), and
bias parameters were initalized to zero.
%
The normalization factor and the update rule for $\theta$
are given by
% XXX Michael: It is crucial, but nowhere mentioned, that the optimization is
% XXX done in an alternating manner. The whole structure relies on this, by
% XXX construction it is vulnerable to optimizing too much towards labels or
% XXX rest at a time. I also identify this as one of the major weak points
% XXX of the method, so it needs to be stated in a way that is not easily
% XXX attackable.
% XXX Maybe we could also talk less about RMSprop.
%
% XXX Danilo: I think we are not alternating, Xrest and Xtask go in
% simultaneously at every bach in every epoch
\begin{eqnarray}
  \begin{split}
    \mathbf{v^{(t+1)}} &= \rho \mathbf{v^{(t+1)}} + (1 - \rho)\left(\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})\right)^2
%v^{(t+1)} = \rho v^{(t+1)} + (1 - \rho)˜|\nabla f
\\
\theta^{(t+1)} &= \theta^{(t)} + \alpha \frac{\nabla_{\theta} f(x^{(t)}, y^{(t)}, \theta^{(t)})}{\sqrt{\mathbf{v^{(t+1)}} + \epsilon}},
  \end{split}
\end{eqnarray}
% XXX Michael: Why, in the formula, do we have two different notations for
% XXX partial derivatives? \partial and \nabla
% XXX Michael: Probably remove the explanations following and just state values
% XXX: Bertrand: what is f ?? Danilo: replaced f by more detailed notation
where $f$ is the loss function computed on a minibatch sample at timestep $t$,
$\alpha$ is the learning rate ($0.00001$),
$\epsilon$ a global damping factor ($10^{-6}$),
and
$\rho$ the decay rate ($0.9$ to deemphasize the magnitude of the gradient).
%
Note that we have also experimented with other solvers
(stochastic gradient descent, adadelta, and adagrad) but found that
\textit{rmsprop} converged faster and with
similar or higher generalization performance.

@@

SS-FLogReg was subsequently trained ($500$ epochs) across parameter choices

@@

higher bias and smaller variance then SS-FLogReg with 100 latent directions.

@@

\begin{table}[h]
  \centering
  \resizebox{1.00\textwidth}{!}{%
\tabcolsep=0.05cm
\begin{tabular}{l|ccccc|ccccc|ccccc}
  \hline
                            & \multicolumn{5}{c|}{\textit{n = 5}}                                                                                                                                                 & \multicolumn{5}{c|}{\textit{n = 20}}                                                                                                                                                      & \multicolumn{5}{c}{\textit{n = 100}}                                                                           \\ \hline
  \textit{}                 & \multicolumn{1}{l}{$\lambda=0$}    & \multicolumn{1}{l}{$\lambda=0.25$} & \multicolumn{1}{l}{$\lambda=0.5$} & \multicolumn{1}{l}{$\lambda=0.75$} & \multicolumn{1}{l|}{$\lambda=1$} & \multicolumn{1}{l}{$\lambda=0$}  & \multicolumn{1}{l}{$\lambda=0.25$}   & \multicolumn{1}{l}{$\lambda=0.5$}   & \multicolumn{1}{l}{$\lambda=0.75$}   & \multicolumn{1}{l|}{$\lambda=1$}   & $\lambda=0$      & $\lambda=0.25$   & $\lambda=0.5$    & $\lambda=0.75$   & \multicolumn{1}{l}{$\lambda=1$}    \\ \hline
  Out-of-sample\\accuracy   & \textit{6,0\%}                     & 88,9\%                             & 95,1\%                            & \textbf{96,5\%}                    & 95,7\%                           & \textit{5,5\%}                   & 97,4\%                               & \textbf{97,8\%}                     & 97,3\%                               & 97,3\%                             & \textit{6,1\%}   & 97,2\%           & 97,0\%           & \textbf{97,8\%}  & 97,4\%                             \\
  Precision (mean)          & \textit{5,9\%}                     & 87,0\%                             & 94,9\%                            & \textbf{96,3\%}                    & 95,4\%                           & \textit{5,1\%}                   & \textbf{97,4\%}                      & 97,1\%                              & 97,0\%                               & 97,0\%                             & \textit{5,9\%}   & 96,9\%           & 96,5\%           & \textbf{97,5\%}  & 96,9\%                             \\
  Recall (mean)             & \textit{5,6\%}                     & 88,3\%                             & 95,2\%                            & \textbf{96,6\%}                    & 95,7\%                           & \textit{4,6\%}                   & \textbf{97,5\%}                      & 97,5\%                              & 97,4\%                               & 97,4\%                             & \textit{7,2\%}   & 97,2\%           & 97,2\%           & \textbf{97,9\%}  & 97,4\%                             \\
  F1 score (mean)           & \textit{4,1\%}                     & 86,6\%                             & 94,9\%                            & \textbf{96,4\%}                    & 95,4\%                           & \textit{3,8\%}                   & \textbf{97,4\%}                      & 97,2\%                              & 97,1\%                               & 97,1\%                             & \textit{5,3\%}   & 97,0\%           & 96,7\%           & \textbf{97,7\%}  & 97,2\%                             \\ \hline
  Out-of-study\\accuracy    & \textit{39,4\%}                    & 60,8\%                             & 54,3\%                            & 60,7\%                             & \textbf{62,9\%}                  & \textit{77,0\%}                  & 79,7\%                               & \textbf{81,9\%}                     & 79,7\%                               & 79,4\%                             & 79,2\%           & \textbf{82,2\%}  & 81,7\%           & 81,3\%           & \textit{75,8\%}                    \\ \hline
  \end{tabular}
}
\vspace{-0.2cm}
 \caption{\textbf{Performance of SS-FLogReg across model parameter choices.}
 Chance is at 5,6\%.
 }
  \label{table_two}
\end{table}

@@

\paragraph{Feature identification.}
We finally examined whether the models
were fit for purpose
(Figure \ref{fig_weights}).
%
To this end, we computed Pearson's correlation between the classifier weights
and the averaged neural activity map for each of the 18 tasks.
%
Ordinary logistic regression thus yielded a mean correlation
of $\rho=0.28$.
%
For SS-FLogReg ($\lambda=0.25, 0.5, 0.75, 1$),
a per-class-weight map was computed by matrix
multiplication of the two inner layers.
Feature identification thus ranged
between $\rho=0.35$ and $\rho=0.55$ for $n=5$,
between $\rho=0.59$ and $\rho=0.69$ for $n=20$, and
between $\rho=0.58$ and $\rho=0.69$ for $n=100$.
%
Consequently,
SS-FLogReg puts higher absolute weights on relevant structure.
This reflects a higher signal-to-noise ratio, in part explained
by the more BOLD-typical local contiguity.
%
Conversely, SS-FLogReg puts lower probability mass on irrelevant structure.
Despite lower interpretability, the
salt-and-pepper-like weight maps obtained from
ordinary logistic regression are sufficient for
good classification performance.
%
\begin{figure}
\begin{centering}
\includegraphics[width=0.98\textwidth]{figures/figure_weights_perc75.png}
\end{centering}
\vspace{-0.1cm}
\caption{\textbf{Classification weight maps}
The voxel predictors corresponding to 5 exemplary
(of 18 total) psychological tasks (\textit{rows})
from the HCP dataset \cite{barch2013}.
\textit{Left column:} multinomial logistic regression (same
implementation but without bottleneck or autoencoder),
\textit{middle column:} semi-supervised factored logistic regression
($n=20$ latent components, $\lambda=0.5$, $\ell_1=0.1$, $\ell_2=0.1$),
\textit{right column:} voxel-wise average of whole-brain
activity maps from each task.
Semi-supervised factored logistic regression
a) puts higher absolute weights on relevant structure,
b) lower ones on irrelevant structure,
and
c) yields BOLD-typical local contiguity (without enforcing a spatial prior).
All values are z-scored and thresholded at the $75^{th}$ percentile.
}
\label{fig_weights}
\end{figure}
%
Hence, SS-FLogReg yielded class weights 
that were much more similar to features of the respective training samples
for any choice of $\lambda$.
SS-FLogReg therefore
captures genuine properties of neural activity patterns,
rather than participant- or study-specific artefacts.
